Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2015-10-17 patchy-selection-paper.tex
moving over to plos

% make supp figures separate
\usepackage{caption}
\DeclareCaptionType{sfigure}[Figure]

@@

%%%%%
% Responses to reviews:
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
% \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review responses file
\newcommand{\responsefile}{patchy-review-responses-second-round.tex}  % name of the review responses file

% override above choices (see makefile)
\ifdef{\submissionnofigs}{  % for submission, no figures
  \reviewresponsesfalse
  \submissionversiontrue
}{
}
\ifdef{\submissionfigs}{  % for reviewers, with figs
  \reviewresponsestrue
  \submissionversionfalse
}{
}
\ifdef{\arxiv}{  % for general use
  \reviewresponsesfalse
  \submissionversionfalse
}{
}

\ifreviewresponses
  \input{review-response-commands}
\else
  \newcommand{\linelabel}[1]{}
  \newcommand{\revpoint}[2]{}
  \newcommand{\includereviews}{}
\fi

@@

These asymptotics fit simulations quite well, as shown in Figure~\ref{fig:sim_occupation_freqs}.
To be clear about the assumptions implicit here,
below we provide a derivation of \eqref{eqn:eqfreq} in section \secref{apx:eqfreq},  \revpoint{3}{5}
as well as justification for the asymptotics below in section \secref{apx:asymptotics}.
In one dimension, the equation can be solved to give $q(x)$ 
in terms of Jacobi elliptic functions; see the Supporting Information.

@@

This expected frequency $q(x)$ is the \emph{time-averaged} occupation frequency,
i.e., the total number of $B$ alleles found across $T$ generations near location $x$, per unit area, divided by $T$.
If, for instance, $q(x)=.01$ and the population density is $\rho=10$ individuals per unit area, 
then in a survey tract of unit area around $x$ we expect to find one individual every 10 generations
-- or, perhaps more likely, 10 individuals every 100 generations.
This points to an important fact: 
close to the original patch, the ``equilibrium frequency'' $q(x)$ describes well the density of the $B$ allele at most times,
but far away from the patch, 
the equilibrium is composed of rare excursions of families of $B$
alleles, interspersed by long periods of absence. 
An example of one of these rare excursions is shown in Figure~\ref{fig:sim_snapshots}.
The relevant time scale on which $B$ alleles migrate between patches is given by the rate of appearance of such families.
% To compute the rate of establishment by such migration,
% we argue that these families (suitably defined) can be approximated by
% independent branching processes whose spatial motion is described by Brownian motion.
% Properties of these, combined with equation \eqref{eqn:eqfreq},
% will lead us eventually to the expression \eqref{eqn:migrate} for the rate of effective migration,
% an approximation that is valid if the distance $R$ is large (at least several times $\sigma/\sqrt{2s_m}$).

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-snapshots}
  \end{center}
  \caption{
  A segment of the temporal dynamics of the same simulation as Figure~\ref{fig:sim_occupation_freqs}B.
  Space is again shown across the horizontal axis, with time on the vertical axis; 
  demes are colored darker red the greater their allele frequency at the corresponding time.
  The 50\% and 5\% frequency contour lines are in grey,
  and the genealogy of 50 individuals in a rare long-distance excursion is traced back with black lines.
  }   \label{fig:sim_snapshots}
\end{figure}

This suggests decomposing the genealogy of $B$ alleles into families in the following way.
First, consider the genealogy of all $B$ alleles that were alive at any time outside of the original patch.  
This is a collection of trees, each rooted at an allele living outside the patch whose parent was born inside the patch.
Next, fix some intermediate distance $r_0$ from the established patch,
and erase from the genealogy every allele that has no ancestors living further away than $r_0$ to the patch.
This fragments the original set of trees into a collection of smaller trees that relate to each other all $B$ alleles living outside of $r_0$ at any time,
and some living inside of $r_0$;
we refer to these trees as ``families''.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated 
by a collection of independent spatial branching processes
whose members are ignored if they enter the original patch,
illustrated in figure~\ref{fig:branching_decomp}.
(This can be made formal in the limit of large population density, also taking $r_0$ large enough that the chance of reaching the original patch is small.)
The opportunity for adaptation depends on how often these families of $B$ alleles encounter the new patch.
Suppose that the area occupied by the new patch is $S$.
We can learn about the rate of arrival of families at $S$ from
the time-averaged number of $B$ alleles expected in $S$ were it not a patch
(i.e., if $B$ was still maladaptive in $S$),
which from equation \eqref{eqn:eqfreq}
is 
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        \rho q(S) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation in new patch per family ) } ,
\end{split}
\end{align}
where $q(S) = \int_S q(y) dy$.
The quantity we wish to compute,
the effective rate at which \emph{families of} migrant $B$ alleles establish in the new patch,
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(S) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in new patch ) } .
    \end{split}
\end{align}
The quantities on the right-hand side depend implicitly on $r_0$,
but their product does not.

%%%% %%%%
\subsubsection{The genealogy of migrant families}
\label{ss:migrant_math}

@@

\begin{figure}[ht!!]
  \begin{center}
    \includegraphics{branching-concept}
  \end{center}
  \caption{
      \textbf{(a)} 
      Cartoon of the decomposition described in the text. 
    Migrants families that get further than $r_0$ from the already-established patch (grey circle)
    are each composed of a ``trunk'' lineage (in red), whose spatial motion we follow,
    embellished by transient branches.
    For the depicted long-lived family, these are further distinguished by those still alive at time $t$ (in black)
    and those that did not survive until $t$ (in grey).
    Figure~\ref{fig:sim_snapshots} has a depiction of such a family, from a simulation. 
    Note that the spatial motions of lineages are independent random walks
    (and hence should be much wigglier than in this cartoon, which confounds space and time).
      \textbf{(b)} 
      Diagram of the area $A'$ used in equation \eqref{eqn:migrate}:
      $R$ is the shortest distance from the old patch to the new,
      and $A'$ is the area of the portion of the new patch that is no more than distance $R+\sigma/\sqrt{2s_m}$
      from the old patch.
} \label{fig:branching_decomp}
\end{figure}

%%%% %%%%
\subsubsection{Heuristics}
\label{ss:heuristics}

@@

To test the robustness of our results to the various approximations we used,
we used individual-based simulations on a one-dimensional lattice of 501 demes,
with $\rho$ haploid individuals per deme,
dispersal to nearby demes with $\sigma=0.95$,
and run for 25,000 generations. 
More details are given below in section \secref{ss:simulations},
and the number of simulations used and parameter values are given in
supplementary tables \ref{stab:mutation_params} and
\ref{stab:migration_params}. 
To estimate the mean time until adaptation by mutation,
we used one centrally located patch of 99 demes
and a mutation rate of $\mu=10^{-5}$,
and to estimate the mean time until adaptation by migration,
we used two centrally located patches of 99 demes
separated by a varying number of demes,
with one patch initialized to have the $B$ allele at frequency 0.8.
In each case, we measured the ``time to adaptation'' as the amount of time until at least 100 $B$ alleles
were present in the previously unadapted patch.
Figure~\ref{fig:sim_times}
summarizes how the results compare to theory,
excluding parameter combinations 
that violate the assumptions discussed above, 
or where a majority of simulations did not adapt by 25,000 generations.
Supplementary figures \ref{sfig:sim_migration_times} and \ref{sfig:sim_mutation_times} 
show all times, 
and depictions of typical simulation runs are shown in supplementary figures
\ref{sfig:sims_1}, \ref{sfig:sims_2}, \ref{sfig:sims_3}, \ref{sfig:sims_4}, \ref{sfig:sims_5}, \ref{sfig:sims_6}, \ref{sfig:sims_7}, and \ref{sfig:sims_8}.

The agreement is reasonable for both.
The theoretical value $1/\mutrate$ underestimates the mean time to mutation
by a factor of around 2 that increases slightly with $s_m$.
This is to be expected for two reasons:
First, we compute the time to reach 100 $B$ alleles, while theory predicts
the time until the progenitor of those 100 $B$ alleles arose.
Second, the expression for $\mutrate$ neglects effects near the boundary of the patch,
and the larger $s_m$ is, 
the harder it is for mutations that arise near the edge of the patch to establish.
The theoretical values $1/\migrate$ again has a margin of error of a factor of about 2.
% with the exception of the weakest disadvantage off the patch, $s_m=10^{-4}$,
% which adapt much more quickly than predicted by theory.
% However, these simulations fall outside the region where we expect the model to do well,
% since $R < \sigma/\sqrt{s_m} \approx 100$.
Supplementary figure \ref{sfig:sims_8} shows simulations at more parameter values.
% since $s_m$ is so small, the second patch is reached by migrants before the migration-selection equilibrium is reached,
% and at migration-selection equilibrium, the two patches act as a single meta-patch.

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      \includegraphics{times-predicted-observed}
  \end{center}
  \caption{
      \textbf{(left)} 
      Time until adaptation by mutation in an isolated patch for different combinations
      of selection coefficient on patches ($s_m$), 
      population density ($\rho$),
      mutation rate ($\mu$),
      and distance between patches ($R$),
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:mutrate} (horizontal axis).
      In all simulations, $s_p=0.01$.
      \textbf{(right)} 
      Time until adaptation by migration from an already occupied patch to a new patch,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:migrate}, with $C=5$ (horizontal axis).
      For each,
      parameter combinations and numbers of simulations are in supplemental tables \ref{stab:mutation_params} and \ref{stab:migration_params};
      shown are the median (points) and interquartile range (lines)
      of the time until the patch had at least 100 adapted individuals,
      for each unique parameter combination.
      Only parameter combinations to which the theory here is applicable were used
      (precisely, if
      $R > \sigma/\sqrt{s_m}$, 
      patch width larger than $\sigma/\sqrt{s_m} \atan(\sqrt{s_m/s_p})$, 
      and $s_m > 2 \sigma \rho$);
      the remaining combinations are shown in figures
      S\ref{sfig:sim_migration_times} and S\ref{sfig:sim_mutation_times}.
  }   \label{fig:sim_times}
\end{figure}

@@

We assume that once a $B$ allele is introduced (by migration or mutation) 
it becomes established in the poor habitat patch rapidly
if it escapes loss by demographic stochasticity.
Under this approximation, and the ``selective exclusion'' just discussed,
after one of the patches becomes colonized by a single $B$ allele, 
the other patch will become colonized by migration or mutation, but not by both. 
As such, the question of how the second patch adapts
simply comes down to whether a new mutation or a migrant allele is the first to become established in the second patch. 
To work out how adaptation is likely to proceed, 
we can compare expressions \eqref{eqn:mutrate} and \eqref{eqn:migrate} above
for the effective rates of adaptation by new mutation and by migration.
We work in one dimension, as the square root term appearing for two dimensions is relatively unimportant.
% (as seen in figure~\ref{fig:sim_occupation_freqs} by the fact that the $d=2$ curve is nearly straight).
\linelabel{rr:sqrt_term}

@@

% One way to visualize this is by a phase diagram.
% In figure \ref{fig:phase_diagram},
% we give a contour plot for the time until adaptation by migration ($\Tmig = 1/\migrate$) and by new mutation ($\Tmut=1/\mutrate$)
% as a function of various parameters. 
% This gives us an idea of not only the parameter space where parallel adaptation is likely ($\Tmig \ll \Tmut$),
% but also where adaptation by any means takes a very long time ($\Tmig$ and $\Tmut$ both very large),
% an important check in applications of these results.
% 
% \begin{figure}[ht]
%   \begin{center}
%     \includegraphics{phase-diagram-log}
%   \end{center}
%   \caption{
%   Shaded contour plots of
%   mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
%   as a function of population density ($\rho$), 
%   distance to an already adapted patch in units of dispersal distance ($R/\sigma$),
%   strength of deleterious selection ($s_m$),
%   and mutation rate scaled by patch area ($\mu A$).
%   The right-hand plot shows that $\migrate$ is relatively insensitive to $s_m$ (thanks to the square root).
%   The black, dotted line shows values where $\Tmig=\Tmut$.
%   Default parameter values were $A=100$, $\rho=10$, $\mu=10^{-8}$, $s_p=0.1$, $s_m=0.01$, $R/\sigma=20$, and $\xi^2=1$.
%   Note that population density $\rho$ and mutation influx $A\mu$ are shown on a log scale.
%   \label{fig:phase_diagram}
%   }
% \end{figure}

@@

We tested this using the same simulations as figure \ref{fig:sim_times},
by using the empirical distributions of the respective times to adaptation
to estimate the probability,
for each parameter combination,
that a new, successful mutation appears 
before a successful migrant arrives from another, already adapted patch.
The results are compared to \eqref{eqn:parallel_prob}
in figure~\ref{fig:sim_probs}.

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      \includegraphics{prob-mutation-compared}
  \end{center}
  \caption{
      Probability that two patches 
      evolve by independent mutations (i.e., convergently),
      as a function of their geographic separation ($R$)
      and the strength of selection between patches ($s_m$).
      The background shades of grey and contours
      give the probability predicted by theory (equation~\eqref{eqn:parallel_prob}, with $C=5$).
      Points correspond to parameter combinations of the simulations in figure \ref{fig:sim_times},
      with the shade of grey and the red numeric label indicating the same probability
      estimated from these simulations (see text for details).
      The two densities, $\rho=100$ and $\rho=1000$ would lie directly on top of each other, so the $\rho=1000$ points have been shifted vertically.
  }   \label{fig:sim_probs}
\end{figure}

%%%%%%%%%%%%
\subsubsection{Multiple patches}

@@

\begin{figure}[ht]
  \begin{center}
      \includegraphics{sim-transit}
  \end{center}
  \caption{
  \textbf{Adaptation by migration:}
  An initially unadapted patch (10 demes at the left end of the range) is colonized around generation 5000;
  a representative lineage is shown in the new patch, which follows the original migrant ``trunk'' lineage
  between the patches and coalesces with a lineage in the unadapted patch.
  In the left panel, as in Figure~\ref{fig:sim_snapshots}, color and contours show allele frequency across space and time;
  other simulation parameters are as in Figures~\ref{fig:sim_occupation_freqs}B and \ref{fig:sim_snapshots}.
  The right panel shows (in red) the extent of the shared haplotype between the two lineages around the selected locus (at position 0),
  which decreases as time progresses. 
  \label{fig:lineagesmotion}
  }
\end{figure}

@@

When an already adapted patch colonizes another through migration,
 the newly colonized patch will inherit a long piece of haplotype around the selected site from the originating patch.
The genetic length of this haplotype is roughly inversely proportional to the
number of generations the allele spends ``in transit'', 
because while very rare, the allele will be mostly in heterozygotes,
and so each generation provides another opportunity for a different haplotype to recombine closer to the transiting $B$ allele.
A large, linked haplotype may still arrive and fix in the new patch,
in which case the haplotype has literally hitchhiked across geography!
Figure~\ref{fig:lineagesmotion} shows a simulation of such an event,
including the lineage that founds an adaptive population on a second patch,
and the length of the haplotype shared between this lineage and one in the original patch.
The time that the lineage is outside the region where the $B$ allele is common (dark red in Figure~\ref{fig:lineagesmotion}), 
the haplotype that accompanies it is broken down rapidly. 
After the lineage establishes on the patch, the rate of decay of the haplotype is slowed significantly, 
since most others with which it recombines have similar haplotypic backgrounds. 
% We first discuss the transit of the lineage, and its accompanying haplotype, and return
% to how haplotypes are whittled down within established patches in the next section.

@@

The mutational target size $\mu$ for the trait is unclear. 
While the Pinacate dark haplotype differs from the light haplotype at four amino acid residues,
it is likely that not all of these changes are needed for a population to begin to adapt. 
Also, there a number of genes besides MC1R at which adaptive changes affecting pigmentation 
have been identified in closely related species and more broadly across vertebrates \citep{Hoekstra:06}.
To span a range of plausible values, we use a low mutation rate of $\mu= 10^{-8}$
(a single base pair),
and a high mutation rate $\mu= 10^{-5}$
(a kilobase).
Finally, we set $A=100\text{km}^2$ (roughly the size of the Pinacate
patch). In
Figure~\ref{fig:mice_prob_parallel} we show the dependence of the probability
of parallel mutation on the distance between lava flow patches using these parameters, 
showing that parallel mutation should become
likely over a scale of tens to a few hundred kilometers between patches.

Given the large selection coefficient associated with the dark allele
on the dark substrate, we expect the initial haplotype associated with
either a new mutation or migrant allele to be large. 
Figure~\ref{fig:mice_prob_parallel} also shows how long the founding haplotype 
shared between populations is expected to be,
from equation \eqref{eqn:haplotype_length}.
The initial length can be quite long between geographically close patches (tens of kilometers). 
However, for the wider cline width ($\sigma/\sqrt{s_m} = 30$km), 
adaptation by migration can still be likely for patches 100km apart, 
but the shared basis may be hard to detect, as the length of shared haplotype can be quite short.

@@

Here we provide additional details regarding the simulations we used
to validate the theory in figures \ref{fig:sim_times} and \ref{fig:sim_probs}.

>>>>

% make supp figures separate
\usepackage{caption}
\DeclareCaptionType{sfigure}[Fig]

@@

%  %%%%%
%  % Responses to reviews:
%  % set this to show line numbers and include responses to reviews or not
%  \newif\ifreviewresponses
%  \reviewresponsestrue  % include them
%  % \reviewresponsesfalse  % don't include them
%  \iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
%  % \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review responses file
%  % \newcommand{\responsefile}{patchy-review-responses-second-round.tex}  % name of the review responses file
%  
%  % override above choices (see makefile)
%  \ifdef{\submissionnofigs}{  % for submission, no figures
%    \reviewresponsesfalse
%    \submissionversiontrue
%  }{
%  }
%  \ifdef{\submissionfigs}{  % for reviewers, with figs
%    \reviewresponsestrue
%    \submissionversionfalse
%  }{
%  }
%  \ifdef{\arxiv}{  % for general use
%    \reviewresponsesfalse
%    \submissionversionfalse
%  }{
%  }
%  
%  \ifreviewresponses
%    \input{review-response-commands}
%  \else
    \newcommand{\linelabel}[1]{}
    \newcommand{\revpoint}[2]{}
    \newcommand{\includereviews}{}
%  \fi

@@

These asymptotics fit simulations quite well, as shown in Fig~\ref{fig:sim_occupation_freqs}.
To be clear about the assumptions implicit here,
below we provide a derivation of \eqref{eqn:eqfreq} in section \secref{apx:eqfreq},  \revpoint{3}{5}
as well as justification for the asymptotics below in section \secref{apx:asymptotics}.
In one dimension, the equation can be solved to give $q(x)$ 
in terms of Jacobi elliptic functions \citep{NEQwiki}; 
see the Supporting Information.

@@

This expected frequency $q(x)$ is the \emph{time-averaged} occupation frequency,
i.e., the total number of $B$ alleles found across $T$ generations near location $x$, per unit area, divided by $T$.
If, for instance, $q(x)=.01$ and the population density is $\rho=10$ individuals per unit area, 
then in a survey tract of unit area around $x$ we expect to find one individual every 10 generations
-- or, perhaps more likely, 10 individuals every 100 generations.
This points to an important fact: 
close to the original patch, the ``equilibrium frequency'' $q(x)$ describes well the density of the $B$ allele at most times,
but far away from the patch, 
the equilibrium is composed of rare excursions of families of $B$
alleles, interspersed by long periods of absence. 
An example of one of these rare excursions is shown in Fig~\ref{fig:sim_snapshots}.
The relevant time scale on which $B$ alleles migrate between patches is given by the rate of appearance of such families.
% To compute the rate of establishment by such migration,
% we argue that these families (suitably defined) can be approximated by
% independent branching processes whose spatial motion is described by Brownian motion.
% Properties of these, combined with equation \eqref{eqn:eqfreq},
% will lead us eventually to the expression \eqref{eqn:migrate} for the rate of effective migration,
% an approximation that is valid if the distance $R$ is large (at least several times $\sigma/\sqrt{2s_m}$).

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-snapshots}
  \end{center}
  \caption{
  A segment of the temporal dynamics of the same simulation as Fig~\ref{fig:sim_occupation_freqs}B.
  Space is again shown across the horizontal axis, with time on the vertical axis; 
  demes are colored darker red the greater their allele frequency at the corresponding time.
  The 50\% and 5\% frequency contour lines are in grey,
  and the genealogy of 50 individuals in a rare long-distance excursion is traced back with black lines.
  }   \label{fig:sim_snapshots}
\end{figure}

This suggests decomposing the genealogy of $B$ alleles into families in the following way.
First, consider the genealogy of all $B$ alleles that were alive at any time outside of the original patch.  
This is a collection of trees, each rooted at an allele living outside the patch whose parent was born inside the patch.
Next, fix some intermediate distance $r_0$ from the established patch,
and erase from the genealogy every allele that has no ancestors living further away than $r_0$ to the patch.
This fragments the original set of trees into a collection of smaller trees that relate to each other all $B$ alleles living outside of $r_0$ at any time,
and some living inside of $r_0$;
we refer to these trees as ``families''.
If $r_0$ is large enough that there are not too many families
and interactions between family members are weak,
then these families can be approximated 
by a collection of independent spatial branching processes
whose members are ignored if they enter the original patch,
illustrated in Fig~\ref{fig:branching_decomp}.
(This can be made formal in the limit of large population density, also taking $r_0$ large enough that the chance of reaching the original patch is small.)
The opportunity for adaptation depends on how often these families of $B$ alleles encounter the new patch.
Suppose that the area occupied by the new patch is $S$.
We can learn about the rate of arrival of families at $S$ from
the time-averaged number of $B$ alleles expected in $S$ were it not a patch
(i.e., if $B$ was still maladaptive in $S$),
which from equation \eqref{eqn:eqfreq}
is 
\begin{align}
    \label{eqn:gestalt_q}
    \begin{split}
        \rho q(S) &= \text{ ( outflux of families ) } \times \text{ ( mean occupation in new patch per family ) } ,
\end{split}
\end{align}
where $q(S) = \int_S q(y) dy$.
The quantity we wish to compute,
the effective rate at which \emph{families of} migrant $B$ alleles establish in the new patch,
is
\begin{align}
    \label{eqn:gestalt_migrate}
    \begin{split}
        \migrate(S) &= \text{ ( outflux of families ) } \times \text{ ( probability a family establishes in new patch ) } .
    \end{split}
\end{align}
The quantities on the right-hand side depend implicitly on $r_0$,
but their product does not.

%%%% %%%%
\paragraph{The genealogy of migrant families}
\label{ss:migrant_math}

@@

\begin{figure}[ht!!]
  \begin{center}
    \includegraphics{branching-concept}
  \end{center}
  \caption{
      \textbf{(a)} 
      Cartoon of the decomposition described in the text. 
    Migrants families that get further than $r_0$ from the already-established patch (grey circle)
    are each composed of a ``trunk'' lineage (in red), whose spatial motion we follow,
    embellished by transient branches.
    For the depicted long-lived family, these are further distinguished by those still alive at time $t$ (in black)
    and those that did not survive until $t$ (in grey).
    Fig~\ref{fig:sim_snapshots} has a depiction of such a family, from a simulation. 
    Note that the spatial motions of lineages are independent random walks
    (and hence should be much wigglier than in this cartoon, which confounds space and time).
      \textbf{(b)} 
      Diagram of the area $A'$ used in equation \eqref{eqn:migrate}:
      $R$ is the shortest distance from the old patch to the new,
      and $A'$ is the area of the portion of the new patch that is no more than distance $R+\sigma/\sqrt{2s_m}$
      from the old patch.
} \label{fig:branching_decomp}
\end{figure}

%%%% %%%%
\paragraph{Heuristics}
\label{ss:heuristics}

@@

To test the robustness of our results to the various approximations we used,
we used individual-based simulations on a one-dimensional lattice of 501 demes,
with $\rho$ haploid individuals per deme,
dispersal to nearby demes with $\sigma=0.95$,
and run for 25,000 generations. 
More details are given below in section \secref{ss:simulations},
and the number of simulations used and parameter values are given in
supplementary tables \ref{stab:mutation_params} and
\ref{stab:migration_params}. 
To estimate the mean time until adaptation by mutation,
we used one centrally located patch of 99 demes
and a mutation rate of $\mu=10^{-5}$,
and to estimate the mean time until adaptation by migration,
we used two centrally located patches of 99 demes
separated by a varying number of demes,
with one patch initialized to have the $B$ allele at frequency 0.8.
In each case, we measured the ``time to adaptation'' as the amount of time until at least 100 $B$ alleles
were present in the previously unadapted patch.
Fig~\ref{fig:sim_times}
summarizes how the results compare to theory,
excluding parameter combinations 
that violate the assumptions discussed above, 
or where a majority of simulations did not adapt by 25,000 generations.
\ref{sfig:sim_migration_times} and \ref{sfig:sim_mutation_times} Fig
show all times, 
and depictions of typical simulation runs are shown in 
\ref{sfig:sims_1}, \ref{sfig:sims_2}, \ref{sfig:sims_3}, \ref{sfig:sims_4}, \ref{sfig:sims_5}, \ref{sfig:sims_6}, \ref{sfig:sims_7}, and \ref{sfig:sims_8} Fig.

The agreement is reasonable for both.
The theoretical value $1/\mutrate$ underestimates the mean time to mutation
by a factor of around 2 that increases slightly with $s_m$.
This is to be expected for two reasons:
First, we compute the time to reach 100 $B$ alleles, while theory predicts
the time until the progenitor of those 100 $B$ alleles arose.
Second, the expression for $\mutrate$ neglects effects near the boundary of the patch,
and the larger $s_m$ is, 
the harder it is for mutations that arise near the edge of the patch to establish.
The theoretical values $1/\migrate$ again has a margin of error of a factor of about 2.
% with the exception of the weakest disadvantage off the patch, $s_m=10^{-4}$,
% which adapt much more quickly than predicted by theory.
% However, these simulations fall outside the region where we expect the model to do well,
% since $R < \sigma/\sqrt{s_m} \approx 100$.
\ref{sfig:sims_8} Fig shows simulations at more parameter values.
% since $s_m$ is so small, the second patch is reached by migrants before the migration-selection equilibrium is reached,
% and at migration-selection equilibrium, the two patches act as a single meta-patch.

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      \includegraphics{times-predicted-observed}
  \end{center}
  \caption{
      \textbf{(left)} 
      Time until adaptation by mutation in an isolated patch for different combinations
      of selection coefficient on patches ($s_m$), 
      population density ($\rho$),
      mutation rate ($\mu$),
      and distance between patches ($R$),
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:mutrate} (horizontal axis).
      In all simulations, $s_p=0.01$.
      \textbf{(right)} 
      Time until adaptation by migration from an already occupied patch to a new patch,
      as observed in simulations (vertical axis),
      and as predicted from equation~\eqref{eqn:migrate}, with $C=5$ (horizontal axis).
      For each,
      parameter combinations and numbers of simulations are in supplemental tables \ref{stab:mutation_params} and \ref{stab:migration_params};
      shown are the median (points) and interquartile range (lines)
      of the time until the patch had at least 100 adapted individuals,
      for each unique parameter combination.
      Only parameter combinations to which the theory here is applicable were used
      (precisely, if
      $R > \sigma/\sqrt{s_m}$, 
      patch width larger than $\sigma/\sqrt{s_m} \atan(\sqrt{s_m/s_p})$, 
      and $s_m > 2 \sigma \rho$);
      the remaining combinations are shown in 
      S\ref{sfig:sim_migration_times} and S\ref{sfig:sim_mutation_times} Fig.
  }   \label{fig:sim_times}
\end{figure}

@@

We assume that once a $B$ allele is introduced (by migration or mutation) 
it becomes established in the poor habitat patch rapidly
if it escapes loss by demographic stochasticity.
Under this approximation, and the ``selective exclusion'' just discussed,
after one of the patches becomes colonized by a single $B$ allele, 
the other patch will become colonized by migration or mutation, but not by both. 
As such, the question of how the second patch adapts
simply comes down to whether a new mutation or a migrant allele is the first to become established in the second patch. 
To work out how adaptation is likely to proceed, 
we can compare expressions \eqref{eqn:mutrate} and \eqref{eqn:migrate} above
for the effective rates of adaptation by new mutation and by migration.
We work in one dimension, as the square root term appearing for two dimensions is relatively unimportant.
% (as seen in Fig~\ref{fig:sim_occupation_freqs} by the fact that the $d=2$ curve is nearly straight).
\linelabel{rr:sqrt_term}

@@

% One way to visualize this is by a phase diagram.
% In Fig \ref{fig:phase_diagram},
% we give a contour plot for the time until adaptation by migration ($\Tmig = 1/\migrate$) and by new mutation ($\Tmut=1/\mutrate$)
% as a function of various parameters. 
% This gives us an idea of not only the parameter space where parallel adaptation is likely ($\Tmig \ll \Tmut$),
% but also where adaptation by any means takes a very long time ($\Tmig$ and $\Tmut$ both very large),
% an important check in applications of these results.
% 
% \begin{figure}[ht]
%   \begin{center}
%     \includegraphics{phase-diagram-log}
%   \end{center}
%   \caption{
%   Shaded contour plots of
%   mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
%   as a function of population density ($\rho$), 
%   distance to an already adapted patch in units of dispersal distance ($R/\sigma$),
%   strength of deleterious selection ($s_m$),
%   and mutation rate scaled by patch area ($\mu A$).
%   The right-hand plot shows that $\migrate$ is relatively insensitive to $s_m$ (thanks to the square root).
%   The black, dotted line shows values where $\Tmig=\Tmut$.
%   Default parameter values were $A=100$, $\rho=10$, $\mu=10^{-8}$, $s_p=0.1$, $s_m=0.01$, $R/\sigma=20$, and $\xi^2=1$.
%   Note that population density $\rho$ and mutation influx $A\mu$ are shown on a log scale.
%   \label{fig:phase_diagram}
%   }
% \end{figure}

@@

We tested this using the same simulations as Fig \ref{fig:sim_times},
by using the empirical distributions of the respective times to adaptation
to estimate the probability,
for each parameter combination,
that a new, successful mutation appears 
before a successful migrant arrives from another, already adapted patch.
The results are compared to \eqref{eqn:parallel_prob}
in Fig~\ref{fig:sim_probs}.

\begin{figure}[ht!]
  \begin{center}
      % made in analyze-sims.R
      \includegraphics{prob-mutation-compared}
  \end{center}
  \caption{
      Probability that two patches 
      evolve by independent mutations (i.e., convergently),
      as a function of their geographic separation ($R$)
      and the strength of selection between patches ($s_m$).
      The background shades of grey and contours
      give the probability predicted by theory (equation~\eqref{eqn:parallel_prob}, with $C=5$).
      Points correspond to parameter combinations of the simulations in Fig \ref{fig:sim_times},
      with the shade of grey and the red numeric label indicating the same probability
      estimated from these simulations (see text for details).
      The two densities, $\rho=100$ and $\rho=1000$ would lie directly on top of each other, so the $\rho=1000$ points have been shifted vertically.
  }   \label{fig:sim_probs}
\end{figure}

%%%%%%%%%%%%
\paragraph{Multiple patches}

@@

\begin{figure}[ht]
  \begin{center}
      \includegraphics{sim-transit}
  \end{center}
  \caption{
  \textbf{Adaptation by migration:}
  An initially unadapted patch (10 demes at the left end of the range) is colonized around generation 5000;
  a representative lineage is shown in the new patch, which follows the original migrant ``trunk'' lineage
  between the patches and coalesces with a lineage in the unadapted patch.
  In the left panel, as in Fig~\ref{fig:sim_snapshots}, color and contours show allele frequency across space and time;
  other simulation parameters are as in Figures~\ref{fig:sim_occupation_freqs}B and \ref{fig:sim_snapshots}.
  The right panel shows (in red) the extent of the shared haplotype between the two lineages around the selected locus (at position 0),
  which decreases as time progresses. 
  \label{fig:lineagesmotion}
  }
\end{figure}

@@

When an already adapted patch colonizes another through migration,
 the newly colonized patch will inherit a long piece of haplotype around the selected site from the originating patch.
The genetic length of this haplotype is roughly inversely proportional to the
number of generations the allele spends ``in transit'', 
because while very rare, the allele will be mostly in heterozygotes,
and so each generation provides another opportunity for a different haplotype to recombine closer to the transiting $B$ allele.
A large, linked haplotype may still arrive and fix in the new patch,
in which case the haplotype has literally hitchhiked across geography!
Fig~\ref{fig:lineagesmotion} shows a simulation of such an event,
including the lineage that founds an adaptive population on a second patch,
and the length of the haplotype shared between this lineage and one in the original patch.
The time that the lineage is outside the region where the $B$ allele is common (dark red in Fig~\ref{fig:lineagesmotion}), 
the haplotype that accompanies it is broken down rapidly. 
After the lineage establishes on the patch, the rate of decay of the haplotype is slowed significantly, 
since most others with which it recombines have similar haplotypic backgrounds. 
% We first discuss the transit of the lineage, and its accompanying haplotype, and return
% to how haplotypes are whittled down within established patches in the next section.

@@

The mutational target size $\mu$ for the trait is unclear. 
While the Pinacate dark haplotype differs from the light haplotype at four amino acid residues,
it is likely that not all of these changes are needed for a population to begin to adapt. 
Also, there a number of genes besides MC1R at which adaptive changes affecting pigmentation 
have been identified in closely related species and more broadly across vertebrates \citep{Hoekstra:06}.
To span a range of plausible values, we use a low mutation rate of $\mu= 10^{-8}$
(a single base pair),
and a high mutation rate $\mu= 10^{-5}$
(a kilobase).
Finally, we set $A=100\text{km}^2$ (roughly the size of the Pinacate
patch). In
Fig~\ref{fig:mice_prob_parallel} we show the dependence of the probability
of parallel mutation on the distance between lava flow patches using these parameters, 
showing that parallel mutation should become
likely over a scale of tens to a few hundred kilometers between patches.

Given the large selection coefficient associated with the dark allele
on the dark substrate, we expect the initial haplotype associated with
either a new mutation or migrant allele to be large. 
Fig~\ref{fig:mice_prob_parallel} also shows how long the founding haplotype 
shared between populations is expected to be,
from equation \eqref{eqn:haplotype_length}.
The initial length can be quite long between geographically close patches (tens of kilometers). 
However, for the wider cline width ($\sigma/\sqrt{s_m} = 30$km), 
adaptation by migration can still be likely for patches 100km apart, 
but the shared basis may be hard to detect, as the length of shared haplotype can be quite short.

@@

In tables \ref{sfig:sim_migration_times}
and \ref{sfig:sim_mutation_times}
we provide additional details regarding the simulations we used
to validate the theory in \ref{fig:sim_times} and \ref{fig:sim_probs} Fig.
